---
title: "Diss data"
output: html_document
date: "2024-03-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the main libraries for the task: 
```{r}
library(readxl)
library(tidyverse)
library(haven)
library(DBI)
```


Importing the ITERATE (International Terrorism: Attributes of Terrorist Events) dataset for 1968 to 2014 and appending the later years up until 2017
**Note: this dataset was provided to me by the LSE in part fulfillment of my MSc thesis (it is thus not available in the repository). The path here is relative to my own directory but all pieces of code relative to the ITERATE dataset will not be replicable unless the user is independently provided with the data.** 
```{r}
data1968 <- read_excel("/dissertation/ITERATE DATA/ITERATE COMMON FILE (Y) 1968-2014 (1).xlsx")
data2015 <- read_excel("/dissertation/ITERATE DATA/ITERATE COMMON FILE (Y) 2015.xls")
data2016 <- read_excel("/dissertation/ITERATE DATA/ITERATE COMMON FILE (Y) 2016.xls")
data2017 <- read_excel("/dissertation/ITERATE DATA/ITERATE COMMON FILE (Y) 2017.xls")
```
```{r}
# FINAL ITERATE COMBINED DATA
# 1968-2017
combined_data <- bind_rows(data1968, data2015, data2016, data2017)
```

## Cleaning ITERATE 
Here, I clean the data to include the organization names which are in a separate file. 
```{r}
orgs <- read_excel("/dissertation/Organizations ITERATE.xlsx",skip = 1)
# make it into a dataframe for easier handling
as.data.frame(orgs)
# rename the header 
colnames(orgs) <- c("col")
# overwrite orgs to be separated into columns
orgs <-separate_wider_delim(orgs, col, delim = " ", names = c("1st grp inititating action", "group"), too_many = "merge")

#remove leading 0's in orgs so that it will match in combined_data
orgs$`1st grp inititating action` <- gsub("^0+", "", orgs$`1st grp inititating action`)

combined_data$`1st grp inititating action` <- as.character(combined_data$`1st grp inititating action`)

# merge the dataset with combined data
combined_data_names <- left_join(combined_data, orgs, by = "1st grp inititating action")
```

# Cleaning group names
```{r}
# Remove content within parentheses and after dash
combined_data_names$group <- gsub("\\s*\\([^\\)]+\\)", "", combined_data_names$group)
combined_data_names$group <- gsub("([A-Z]), ([^A-Z\\s]*$)", "", combined_data_names$group) 


# Remove leading and trailing spaces
combined_data_names$group <- trimws(combined_data_names$group)
```


# Loading the BAAD dataset 1998-2004
```{r}
# this one can be left as is 
# 1998-2004
baad <- read_dta("/dissertation/BAAD2.dta")

baad <- baad %>%
	rename("Year" = "year") %>%
	rename("group" = "org")

# do the same regex for the baad
baad$group <- gsub("\\s*\\([^\\)]+\\)", "", baad$group)
# add a regex for things after a comma
baad$group <- gsub(",[^,]*", "", baad$group)
```

# cleaning the orgs data
```{r}
# use regex to clean differences in nomenclature
orgs$group <- gsub("\\s*\\([^\\)]+\\)", "", orgs$group)
orgs$group <- gsub("\\s*[–‑-]\\s*[^\\s]+", "", orgs$group) 
orgs$group <- gsub("([A-Z]), ([^A-Z\\s]*$)", "", orgs$group) 
orgs$group <- trimws(orgs$group)
```

```{r}
# Iterate through each organization name in baad
for (i in 1:nrow(baad)) {
  # Get the organization name
  org_name <- baad$group[i]
  baad$org_code <- baad$org_code
  # Initialize code as NA
  code <- NA
  
  # Iterate through each organization name in orgs
  for (j in 1:nrow(orgs)) {
    # Get the organization name and code
    org_name_pattern <- orgs$group[j]
    org_code <- orgs$`1st grp inititating action`[j]
    
    # Check if org_name matches org_name_pattern using regular expressions
    if (grepl(org_name_pattern, org_name, ignore.case = TRUE)) {
      # If a match is found, assign the corresponding code
      code <- org_code
      break  # Exit the loop if a match is found
    }
  }
  
  # Add the code to baad dataset
  baad$org_code[i] <- code
}

# manually assign value for missing organization to its equivalent
baad$org_code[baad$group == "Sudan Revolutionary Front"] <- 6251
```


```{r, Rename some elements and clean further}
combined_data_names<- combined_data_names %>% 
	select(-`Terrorists wounded`, 
				 -`Terrorists Wounded`, 
				 -`Foreign wounded`, 
				 -`US Wounded`, 
				 -`Tpye of Weapon used`, 
				 -`...44`) %>%
	mutate(`# of nationalities` = ifelse(`# of nationalities` == 9 | `# of nationalities` == 99, NA, `# of nationalities`)) %>%
	mutate(`Total # of nationalities` = ifelse(`Total # of nationalities` == 99, NA, `Total # of nationalities`)) %>%
	mutate(`Type of state sponsorship` = ifelse(`Type of state sponsorship` == 9, NA, `Type of state sponsorship`)) %>%
	mutate(`# of terrorist grps directly involved`= ifelse(`# of terrorist grps directly involved` == 9, NA, `# of terrorist grps directly involved`)) %>% 
	mutate(`# of terrorists in attack force`= ifelse(`# of terrorists in attack force` == 99, NA, `# of terrorists in attack force`)) %>% 
	mutate(`# of female terrorists`= ifelse(`# of female terrorists` == 99, NA, `# of female terrorists`)) %>% 
	mutate("Recidivists in attack force" = ifelse("Recidivists in attack force" == 99, NA, `Recidivists in attack force`)) %>% 
	mutate(`# of victims` = ifelse(`# of victims` == 999, NA, `# of victims`)) %>% 
	mutate(`Greatest Number of Individuals Wounded` = ifelse(`Greatest Number of Individuals Wounded` == 999, NA, `Greatest Number of Individuals Wounded`)) 

combined_data_names <- combined_data_names %>% mutate(`group`= ifelse(`group`== "UNKNOWN/", "unknown", `group`))
```
```{r getting the index (not part of data cleaning)}
# Print column index and corresponding column names
for (i in seq_along(names(combined_data_names))) {
  print(paste("Column index:", i, "Column name:", names(combined_data_names)[i]))
}
```


# Extended Data on Terrorist Groups
```{r adding the new dataset to the existing dataset}
combo_extend <- read_excel("/dissertation/org_list.xlsx")
combo_extend <- combo_extend %>% select(-`...2`, -`...3`,  -`...4`, -`...5`, -`...6`, -`...7`, -`...8`, -`...9`,)
combo_extend <- as.data.frame(combo_extend)

# Rename the header 
colnames(combo_extend) <- c("extended_name", "col")

# Separate the second column into two columns, allowing for extra merges
combo_extend <- separate(combo_extend, col, into = c("id", "org_name"), extra = "merge") 

# Check if there is an extra column that was created erroneously
combo_extend <- combo_extend[, c("extended_name", "id", "org_name")]

# Clean out the brackets from both columns
combo_extend$org_name <- gsub("\\s*\\([^\\)]+\\)", "", combo_extend$org_name)
combo_extend$extended_name <- gsub("\\s*\\([^\\)]+\\)", "", combo_extend$extended_name)

# remove apostrophes
combo_extend$extended_name <- gsub('"', "", combo_extend$extended_name)
combo_extend$org_name <- gsub('"', "", combo_extend$org_name)

# remove square brackets
combo_extend$extended_name <- gsub("\\[.*?\\]", "", combo_extend$extended_name)
combo_extend$extended_name <- trimws(combo_extend$extended_name)
```

```{r}
# Initialize extend_id column with NA
combo_extend$extend_id <- NA

# Iterate through each row in combo_extend
for (i in 1:nrow(combo_extend)) {
  # Get the value from column 1
  value1 <- combo_extend$extended_name[i]
  
  # Initialize code as NA
  code <- NA
  
  # Iterate through each row in combo_extend to find matching value in column 3
  for (j in 1:nrow(combo_extend)) {
    # Get the value from column 3
    value3 <- combo_extend$org_name[j]
    
    # Check if value1 matches any part of value3 using regex
    if (!is.na(value1) && !is.na(value3) && grepl(value1, value3, ignore.case = TRUE)) {
      # If a match is found, assign the corresponding code from column 2
      code <- combo_extend$id[j]
      break  # Exit the loop if a match is found
    }
  }
  
  # Assign the code to extend_id
  combo_extend$extend_id[i] <- code
}

combo_extend$extend_id[combo_extend$extended_name == "23rd of September Communist League"] <- 0703
combo_extend$extend_id[combo_extend$extended_name == "2nd of June Movement"] <- 2552

#remove leading 0's
combo_extend$extend_id <- gsub("^0+", "", combo_extend$extend_id)

combo_extend_selected <- combo_extend %>% 
	select(-org_name, -id)%>% 
	rename(gname = extended_name)

```

```{r add the ID info back into the other dataframe}
extend <- read_excel("/dissertation/EDTG_Data.xls")
extend <- extend %>% select(-gid)
# re-clean the elements in the other table to make it fit 
extend$gname <- gsub("\\s*\\([^\\)]+\\)", "", extend$gname)

# left join 
extended_terror <- left_join(extend, combo_extend_selected, by = "gname")

extended_terror <- extended_terror %>% 
	rename(Year = year)
```


# Building a SQL-friendly Database
```{r}
group_db <- DBI::dbConnect(RSQLite::SQLite(), "group_db.db")
dbWriteTable(group_db, "combined_data_names", combined_data_names)
dbWriteTable(group_db, "baad", baad)
dbWriteTable(group_db, "extended_terror", extended_terror)
```

```{r}
extended_terror <- extended_terror %>% 
	rename(group = gname)
# Assuming df1 and df2 are your two data frames
full_combined <- merge(baad, extended_terror, by = c("group", "Year"), all = TRUE)
full_combined <- full_combined %>% 
	select(-c(torg, torg_year, hbccode, hbiso, hb_iso_cc))

dbWriteTable(group_db, "final_data", full_combined)
```




