---
title: "data imputation"
output: html_document
date: "2024-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
library(DBI)
library(tidyverse)
library(glmnet)
library(randomForest)
library(mice)
library(VIM)
library(countrycode)
library(caret)
library(missForest)
```

## Loading Data & Preliminary Cleaning of Final Dataset (pre-imputation)
Data stored in the relational database
```{r}
group_db <- DBI::dbConnect(RSQLite::SQLite(), "group_db.db")
final_data <- dbGetQuery(group_db, "
												 SELECT *
												 FROM final_data;")

```

```{rRemoving post variables}
final_data <- final_data %>% 
	select(-c(total_atks, total_casualties, nonterr_casualties, terr_POP, diversity, shr_trans, peak_size, diversity))
```


## Cleaning data from redundancies 

```{r, reconciling the values from different datasets on the same variables} 
# using max because we are interested in all observed behavior 

# drug trafficking 
final_data$combined_drug_trafficking <- pmax(final_data$drugtk, final_data$fddrugtk, na.rm = TRUE)
final_data$combined_drug_trafficking <- as.factor(final_data$combined_drug_trafficking)
levels(final_data$combined_drug_trafficking) <- c(0,1)

# kidnapping
final_data$combined_kidnap <- pmax(final_data$kidnap, final_data$fdkidnap, na.rm = TRUE)
final_data$combined_kidnap <- as.factor(final_data$combined_kidnap)
levels(final_data$combined_kidnap) <- c(0,1)

# state sponsorship
final_data$combined_state_sponsor <- pmax(final_data$state_sponsor, final_data$fdstate, na.rm = TRUE)
final_data$combined_state_sponsor <- as.factor(final_data$combined_state_sponsor)
levels(final_data$combined_state_sponsor) <- c(0,1)

# existence of hierarchy
final_data$combined_hierarch <- pmax(final_data$lead_hierarch.x, final_data$lead_hierarch.y, na.rm = TRUE)
final_data$combined_hierarch <- as.factor(final_data$combined_hierarch)
levels(final_data$combined_hierarch) <- c(0,1)

# existence of leftist ideology
final_data$combined_left <- pmax(final_data$left.x, final_data$left.y, na.rm = TRUE)
final_data$combined_left <- as.factor(final_data$combined_left)
levels(final_data$combined_left) <- c(0,1)

# existence of religious ideology 
final_data$combined_reli <- pmax(final_data$reli, final_data$rel, na.rm = TRUE)
final_data$combined_reli <- as.factor(final_data$combined_reli)
levels(final_data$combined_reli) <- c(0,1)

# group controls land (yes, no)
final_data$combined_terrctrl <- pmax(final_data$terrcntrl, final_data$terrctrl, na.rm = TRUE)
final_data$combined_terrctrl <- as.factor(final_data$combined_terrctrl)
levels(final_data$combined_terrctrl) <- c(0,1)

# number of deaths caused by the organization in a given year
final_data$combined_deaths <- pmax(final_data$ucdpbd, final_data$total_deaths, na.rm = TRUE)

# social services
# if 9 then NA and then pmax
final_data <- final_data %>%
  mutate(social_service = ifelse(social_service == 9, NA, social_service))
final_data$soc_service <- pmax(final_data$socsvcs, final_data$social_service, na.rm = TRUE)

# duration/organization age
final_data$combined_duration <- pmax(final_data$age, final_data$duration, na.rm = TRUE)
```

```{r Reconciling and combining information about the base of the organization}
# Function to merge 'hbase' and 'base' columns
merge_bases <- function(hbase, base) {
  # Split the 'base' column by comma and convert to list of vectors
  base_split <- strsplit(base, ",\\s*")
  
  # Combine 'hbase' and 'base_split' into a single list of vectors
  combined <- mapply(function(h, b) {
    unique(na.omit(c(h, unlist(b))))
  }, hbase, base_split, SIMPLIFY = FALSE)
  
  # Convert list of vectors to a single string with comma separation
  combined <- sapply(combined, function(x) {
    if (length(x) == 0) return(NA)
    paste(x, collapse = ", ")
  })
  
  return(combined)
}

# Apply the function to the dataset
final_data$combined_base <- merge_bases(final_data$hbase, final_data$base)
```

```{r BASE COUNTS}
# Function to count the number of countries in combined_base
count_bases <- function(base_str) {
	return(length(strsplit(base_str, ",\\s*")[[1]]))
}

# Apply the function to create num_bases column
final_data$num_base <- sapply(final_data$combined_base, count_bases)

base_counts <- final_data %>% select(num_base, combined_base )
```

```{r Binary multiple bases}
final_data <- final_data %>%
  mutate(mul_bases = ifelse(num_base > 1, 1, 0))
```

## Region level-variables
Homogenizing the region variables and codes in the dataset 
Missingness can here be clearly dealt with before imputation by process of using the region names to fill in the binary variables

```{r loading package for regional belonging}
# Get all country names
all_countries <- countrycode::codelist$country.name.en

# Get the region for each country
regions <- countrycode(all_countries, "country.name", "region")

# Combine country names and regions into a data frame
country_region_df <- data.frame(country = all_countries, region = regions)
```
```{r EAST ASIA AND PACIFIC}
# Filter countries in the East Asia & Pacific region
eap_countries <- country_region_df %>%
  filter(region == "East Asia & Pacific") %>%
  pull(country)
# Function to check if any country in combined_base is in the EAP region
is_eap <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% eap_countries)
}

# Update EAP based on combined_base
final_data <- final_data %>%
  mutate(EAP = ifelse(is.na(EAP), sapply(combined_base, is_eap), EAP))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(EAP = ifelse(EAP, 1, 0))
```

```{r EUROPE AND CENTRAL ASIA}
eca_countries <- country_region_df %>%
  filter(region == "Europe & Central Asia") %>%
  pull(country)

is_eca <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% eca_countries)
}

# Update ECA based on combined_base
final_data <- final_data %>%
  mutate(ECA = ifelse(is.na(ECA), sapply(combined_base, is_eca), ECA))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(ECA = ifelse(ECA, 1, 0))
```

```{r LATIN AMERICA AND CARRIBEAN}
lac_countries <- country_region_df %>%
  filter(region == "Latin America & Caribbean") %>%
  pull(country)

is_lac <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% lac_countries)
}

# Update LAC based on combined_base
final_data <- final_data %>%
  mutate(LAC = ifelse(is.na(LAC), sapply(combined_base, is_lac), LAC))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(LAC = ifelse(LAC, 1, 0))
```

```{r MIDDLE EAST AND NORTH AFRICA}
mena_countries <- country_region_df %>%
  filter(region == "Middle East & North Africa") %>%
  pull(country)

is_mena <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% mena_countries)
}

# Update MENA based on combined_base
final_data <- final_data %>%
  mutate(MENA = ifelse(is.na(MENA), sapply(combined_base, is_mena), MENA))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(MENA = ifelse(MENA, 1, 0))
```

```{r NORTH AMERICA}
na_countries <- country_region_df %>%
  filter(region == "North America") %>%
  pull(country)

final_data <- final_data %>% 
	rename(North_America = "NA")

is_na <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% na_countries)
}

# Update NA based on combined_base
final_data <- final_data %>%
  mutate(North_America = ifelse(is.na(North_America), sapply(combined_base, is_na), North_America))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(North_America = ifelse(North_America, 1, 0))
```

```{r SOUTH ASIA}
sas_countries <- country_region_df %>%
  filter(region == "South Asia") %>%
  pull(country)

is_sas <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% sas_countries)
}

# Update SAS based on combined_base
final_data <- final_data %>%
  mutate(SAS = ifelse(is.na(SAS), sapply(combined_base, is_sas), SAS))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(SAS = ifelse(SAS, 1, 0))
```

```{r SUB-SAHARAN AFRICA}
ssa_countries <- country_region_df %>%
  filter(region == "Sub-Saharan Africa") %>%
  pull(country)

is_ssa <- function(base) {
  if (is.na(base)) return(NA)
  countries <- strsplit(base, ",\\s*")[[1]]
  any(countries %in% ssa_countries)
}

# Update SSA based on combined_base
final_data <- final_data %>%
  mutate(SSA = ifelse(is.na(SSA), sapply(combined_base, is_ssa), SSA))

# Convert logical to binary (1 for TRUE, 0 for FALSE)
final_data <- final_data %>%
  mutate(SSA = ifelse(SSA, 1, 0))
```

## Size 
```{r reconciling the org. size variables}
# re-organizing the levels 
final_data$size <- ifelse(final_data$size == 1, 1, final_data$size)
final_data$size <- ifelse(final_data$size == 10, 2, final_data$size)
final_data$size <- ifelse(final_data$size == 100, 3, final_data$size)
final_data$size <- ifelse(final_data$size == 1000, 4, final_data$size)
final_data$size <- ifelse(final_data$size == 10000, 5, final_data$size)
final_data$combined_size <- pmax(final_data$size, final_data$size_rec, na.rm = TRUE)
final_data$combined_size <- as.factor(final_data$combined_size)
levels(final_data$combined_size) <- c(1, 2, 3, 4, 5)
```

```{r}
# removing redundant variables from the original dataset 
final_data <- final_data %>% 
	select(-c(drugtk, fddrugtk, kidnap, fdkidnap, state_sponsor, 
						fdstate, lead_hierarch.x, lead_hierarch.y, left.x, 
						left.y, reli, rel, extend_id, terrcntrl, terrctrl, 
						ucdpbd, total_deaths, 
						size, size_rec, org_code, socsvcs, social_service, 
						age, duration, base, hbase, 
						terr_deaths, terr_injuries, nonterr_deaths, terr_casualties, 
						nonterr_injuries, total_injuries, combined_deaths))

```



## Crime Variables

```{r reduce missingness on group-identity by using other ideology information}
# function to update the values in the combined_left variable to reflect the contents of the other columns
# left wing ideology 
update_combined_left <- function(df) {
  df %>%
    mutate(combined_left = ifelse(is.na(combined_left) & (right == 1 | combined_reli == 1 | nat == 1), 
                                  0, combined_left))
}

# Reducing the missingness in those columns
final_data <- update_combined_left(final_data)

# right wing ideology 
update_right <- function(df){
	df %>% 
		mutate(right = ifelse(is.na(right) & (combined_left == 1 | combined_reli == 1 | nat == 1),
													0, right))
}
# reduce the missingness
final_data <- update_right(final_data)


# religious ideology 
update_relig <- function(df){
	df %>% 
		mutate(combined_reli = ifelse(is.na(combined_reli) & (combined_left == 1 | right == 1 | nat == 1), 
																		 0, combined_reli))
}
final_data <- update_relig(final_data)

# nationalism
update_nat <- function(df){
	df %>% 
		mutate(nat = ifelse(is.na(nat) & (combined_left == 1 | right == 1 | combined_reli == 1), 
												0, nat))
}
final_data <- update_nat(final_data)

# territorial change
update_terr <- function(df){
	df %>% 
		mutate(tch = ifelse(is.na(tch) & (nat == 1), 
												1, tch)) %>% 
		mutate(tch = ifelse(is.na(tch) & (sq == 1), 
												0, tch))
}
final_data <- update_terr(final_data)

# update status quo 
update_sq <- function(df){
	df %>% 
		mutate(sq = ifelse(is.na(sq) & (pch == 1 | ercsr == 1 | tch == 1), 
											 0, sq)) %>% 
		mutate(sq = ifelse(is.na(sq) & (pch == 0 & ercsr == 0 & tch == 0), 
											 1, sq))
}
final_data <- update_sq(final_data)

# policy change
update_pch <- function(df){
	df %>% 
		mutate(pch = ifelse(is.na(pch) & (sq == 1), 0, pch))
}
final_data <- update_pch(final_data)

# regime change
update_ercsr <- function(df){
	df %>% 
		mutate(ercsr = ifelse(is.na(ercsr) & (sq == 1), 0, ercsr))
}
final_data <- update_ercsr(final_data)
```

```{r reducing crime missingness by assuming unobserved crime = no crime}
# we can assume that if it has been unobserved it didn't happen, just like Asal et al. 2016 do for their original dataset 
final_data <- final_data %>% 
	mutate(combined_drug_trafficking = ifelse(is.na(combined_drug_trafficking), 0, combined_drug_trafficking)) %>% 
	mutate(combined_kidnap = ifelse(is.na(combined_kidnap), 0, combined_kidnap)) %>% 
	mutate(fdsmuggl = ifelse(is.na(fdsmuggl), 0, fdsmuggl)) %>% 
	mutate(fdrob = ifelse(is.na(fdrob), 0, fdrob)) %>% 
	mutate(fdextort = ifelse(is.na(fdextort), 0, fdextort)) %>% 
	mutate(combined_state_sponsor = ifelse(is.na(combined_state_sponsor), 0, combined_state_sponsor))

final_data <- final_data %>% 
	mutate(combined_terrctrl = ifelse(is.na(combined_terrctrl), 0, combined_terrctrl))
```


```{r re-conciling the social services variables}
# amending social control: if a group  == 1 in "social_service" then it should be == 1 in public service
final_data <- final_data %>%
  group_by(group) %>%
  mutate(
    public_service = case_when(
      soc_service == 1 ~ 1,  # Step 1: If social_service is 1, public_service should be 1
      any(soc_service == 1) ~ 1
    )
  ) %>%
  ungroup() 
```

```{r}
final_data <- final_data %>%
  mutate(fdextort = factor(fdextort)) %>% 
	mutate(fdsmuggl = factor(fdsmuggl)) %>% 
	mutate(fdrob = factor(fdrob)) %>% 
	mutate(end = factor(end)) %>% 
	mutate(ethn = factor(ethn)) %>% 
	mutate(nat = factor(nat)) %>% 
	mutate(right = factor(right)) %>% 
	mutate(ercsr = factor(ercsr)) %>% 
	mutate(pch = factor(pch)) %>% 
	mutate(sq = factor(sq)) %>% 
	mutate(tch = factor(tch)) %>% 
	mutate(crim_degr_py = as.numeric(crim_degr_py)) %>% 
	mutate(num_base = as.numeric(num_base)) %>% 
	mutate(endstr = factor(endstr)) %>% 
	mutate(public_service = factor(public_service)) %>% 
	mutate(fate_leader = factor(fate_leader)) %>% 
	mutate(mul_bases = factor(mul_bases)) %>% 
	mutate(soc_service = factor(soc_service)) %>% 
	mutate(EAP = factor(EAP)) %>% 
	mutate(ECA = factor(ECA)) %>% 
	mutate(LAC = factor(LAC)) %>% 
	mutate(MENA = factor(MENA)) %>% 
	mutate(North_America = factor(North_America)) %>% 
	mutate(SAS = factor(SAS)) %>% 
	mutate(SSA = factor(SSA)) %>% 
	mutate(num_sponsor = as.numeric(num_sponsor)) %>% 
	mutate(num_allies = as.numeric(num_allies)) %>% 
	mutate(num_rivals = as.numeric(num_rivals)) %>% 
	mutate(num_enemies = as.numeric(num_enemies)) %>% 
	mutate(combined_size = factor(combined_size)) %>% 
  mutate(combined_reli = factor(combined_reli, levels = c(0,1))) %>% 
  mutate(combined_left = factor(combined_left, levels = c(0,1))) %>% 
  mutate(combined_hierarch = factor(combined_hierarch)) %>% 
  mutate(combined_drug_trafficking = factor(combined_drug_trafficking, levels = c(0,1))) %>%
  mutate(combined_kidnap = factor(combined_kidnap, levels = c(0,1))) %>%
  mutate(combined_state_sponsor = factor(combined_state_sponsor, levels = c(0,1))) %>%
  mutate(combined_terrctrl = factor(combined_terrctrl, levels = c(0,1)))


# removing columns with too many levels for imputation
# these have no missing data
final_data_imp <- final_data %>% 
	select(-group, - combined_base) 

final_data_imp <- as.data.frame(final_data_imp)
```

## Imputation
Imputing using missForest and then saving file as a CSV
```{r, warning = FALSE, message=FALSE}
control_params <- trainControl(method = "cv", number = 5)

# Define mtry options
mtry_grid <- expand.grid(mtry = c(2, 5, sqrt(ncol(final_data_imp)), 10))

# Train missForest models with different mtry values using cross-validation
missForest_models <- lapply(mtry_grid$mtry, function(mtry) {
  missForest(final_data_imp, maxiter = 10, ntree = 100, replace = TRUE, mtry = mtry)
})

# Extract the OOBerror (NRMSE) from each run
nrmse_values <- sapply(missForest_models, function(model) model$OOBerror[1])

# Find the optimal mtry value
optimal_mtry <- mtry_grid$mtry[which.min(nrmse_values)]

# Train missForest model with the optimal mtry value
optimal_missForest_model <- missForest(final_data_imp, maxiter = 10, ntree = 100, replace = TRUE, mtry = optimal_mtry)

# Impute missing values in the original dataset using the trained model
imputed_data_train <- optimal_missForest_model$ximp

# evaluating performance
optimal_oob_error <- mean(optimal_missForest_model$OOBerror)

# Print the perf
cat("Optimal model's OOB error is:", optimal_oob_error)
```

```{r binding back the year and the combined base into the dataframe}
cols_to_add <- final_data %>% 
	select(group, combined_base)
final_data_complete <- cbind(cols_to_add, imputed_data_train)
str(final_data_complete)
```

```{r no need to run this}
final_data_complete <- final_data_complete %>%
  mutate(fdextort = factor(fdextort, levels = c(0,1))) %>% 
	mutate(fdsmuggl = factor(fdsmuggl, levels = c(0,1))) %>% 
	mutate(fdrob = factor(fdrob, levels = c(0,1))) %>% 
	mutate(end = factor(end)) %>% 
	mutate(ethn = factor(ethn)) %>% 
	mutate(nat = factor(nat)) %>% 
	mutate(right = factor(right)) %>% 
	mutate(ercsr = factor(ercsr)) %>% 
	mutate(pch = factor(pch)) %>% 
	mutate(sq = factor(sq)) %>% 
	mutate(tch = factor(tch)) %>% 
	mutate(crim_degr_py = as.numeric(crim_degr_py)) %>% 
	mutate(num_base = as.numeric(num_base)) %>% 
	mutate(endstr = factor(endstr)) %>% 
	mutate(public_service = factor(public_service)) %>% 
	mutate(fate_leader = factor(fate_leader)) %>% 
	mutate(mul_bases = factor(mul_bases)) %>% 
	mutate(soc_service = factor(soc_service)) %>% 
	mutate(EAP = factor(EAP)) %>% 
	mutate(ECA = factor(ECA)) %>% 
	mutate(LAC = factor(LAC)) %>% 
	mutate(MENA = factor(MENA)) %>% 
	mutate(North_America = factor(North_America)) %>% 
	mutate(SAS = factor(SAS)) %>% 
	mutate(SSA = factor(SSA)) %>% 
	mutate(num_sponsor = as.numeric(num_sponsor)) %>% 
	mutate(num_allies = as.numeric(num_allies)) %>% 
	mutate(num_rivals = as.numeric(num_rivals)) %>% 
	mutate(num_enemies = as.numeric(num_enemies)) %>% 
	mutate(combined_size = factor(combined_size)) %>% 
  mutate(combined_reli = as.factor(combined_reli)) %>% 
  mutate(combined_left = as.factor(combined_left)) %>% 
  mutate(combined_hierarch = as.factor(combined_hierarch)) %>% 
  mutate(combined_drug_trafficking = as.factor(combined_drug_trafficking)) %>%
  mutate(combined_kidnap = as.factor(combined_kidnap)) %>%
  mutate(combined_state_sponsor = as.factor(combined_state_sponsor)) %>%
  mutate(combined_terrctrl = as.factor(combined_terrctrl))
```

```{r save to csv}
write.csv(final_data_complete, "imputed_cleaned_data.csv")
````




